---
title: "Beyond the Label Price Competition and Promotional Effectiveness in Mouthwash"
author: "Jeff Clancy"
date: "`r Sys.Date()`"
output: html_document
---

# Setup

## Packages

```{r}
library(readxl)
library(dplyr)
library(tidyr)
library(purrr)
library(broom)
library(ggplot2)
library(ggrepel)
library(tibble)
library(scales)
```

## Read in data

```{r}
transactions <- read_excel("dunnhumby - Breakfast at the Frat.xlsx", sheet = "dh Transaction Data", skip = 1)
products <- read_excel("dunnhumby - Breakfast at the Frat.xlsx", sheet = "dh Products Lookup", skip = 1)
```

## Merge & Filter

```{r}
data <- merge(transactions, products, by = "UPC")
data <- subset(data, CATEGORY == "ORAL HYGIENE PRODUCTS")
data <- subset(data, DESCRIPTION != "COLG SPEARMINT WISP") # Removing this product as it is not mouthwash
data$DESCRIPTION <- paste(data$DESCRIPTION, data$PRODUCT_SIZE, sep = " - ")

head(data)
```

# EDA

## Product Level Totals

```{r}
mw_total <- data %>%
  group_by(MANUFACTURER, PRODUCT_SIZE, DESCRIPTION) %>%
  summarize(Sales = sum(SPEND, na.rm = TRUE),
            Units_Sold = sum(UNITS, na.rm = TRUE),
            Base_Price = round(mean(BASE_PRICE, na.rm = TRUE), 2),
            .groups = "drop") %>%
  mutate(across(c(DESCRIPTION, PRODUCT_SIZE, MANUFACTURER), as.factor))

print(mw_total)
```

## Plotting Sales by Brand

```{r, fig.height=5, fig.width=8}
ggplot(mw_total %>% arrange(desc(Sales)),
       aes(x = reorder(DESCRIPTION, Sales), y = Sales, fill = MANUFACTURER)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0("$", round(Sales / 1000), "K")), 
            hjust = -0.1, size = 3) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0("$", round(x / 1000), "K"),
    limits = c(0, 350000)  # Set y-axis max to $350K
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "right"  # keep legend on right
  ) +
  labs(
    title = "Mouthwash Sales by Brand & Product", 
    x = "Product", 
    y = "Sales"
  )
```

## Plotting Units by Brand

```{r, fig.height=5, fig.width=8}
ggplot(mw_total %>% arrange(desc(Units_Sold)),
       aes(x = reorder(DESCRIPTION, Units_Sold), y = Units_Sold, fill = MANUFACTURER)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Units_Sold / 1000, 1), "K")), hjust = -0.1, size = 3) +
  coord_flip() +
  theme_minimal() +
  theme(panel.grid = element_blank()) +
  labs(title = "Units Sold by Product", x = "Product", y = "Units Sold") +
  ylim(0, max(mw_total$Units_Sold) * 1.1)
```

# Modeling

## Setting up the data for modeling

```{r}
brand_lookup <- data %>% distinct(DESCRIPTION) %>% arrange(DESCRIPTION) %>% mutate(brand_id = row_number())
data <- data %>% left_join(brand_lookup, by = "DESCRIPTION")

wide_data <- data %>%
  group_by(WEEK_END_DATE, brand_id) %>%
  summarize(
    MOVE = sum(UNITS, na.rm = TRUE),
    LPRICE = log(mean(PRICE, na.rm = TRUE)),
    FEATURE = max(FEATURE, na.rm = TRUE),
    DISPLAY = max(DISPLAY, na.rm = TRUE),
    TPR = max(TPR_ONLY, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_wider(
    names_from = brand_id,
    values_from = c(MOVE, LPRICE, FEATURE, DISPLAY, TPR),
    names_sep = ""
  )
```

## Log Log Model Using Sales with Backwards Selection

```{r}
sales_model <- function(brand_num, data = wide_data) {
  move_var <- paste0("MOVE", brand_num)
  price_var <- paste0("LPRICE", brand_num)
  feature_var <- paste0("FEATURE", brand_num)
  display_var <- paste0("DISPLAY", brand_num)
  tpr_var <- paste0("TPR", brand_num)

  df <- data %>%
    filter(!is.na(.data[[move_var]]), .data[[move_var]] > 0) %>%
    mutate(
      log_sales = log(.data[[move_var]]),
      log_price = .data[[price_var]],
      feature = .data[[feature_var]],
      display = .data[[display_var]],
      tpr = .data[[tpr_var]]
    )

  for (i in 1:max(brand_lookup$brand_id)) {
    if (i != brand_num) {
      df[[paste0("log_price", i)]]  <- data[[paste0("LPRICE", i)]][match(rownames(df), rownames(data))]
      df[[paste0("feature", i)]]    <- data[[paste0("FEATURE", i)]][match(rownames(df), rownames(data))]
      df[[paste0("display", i)]]    <- data[[paste0("DISPLAY", i)]][match(rownames(df), rownames(data))]
      df[[paste0("tpr", i)]]        <- data[[paste0("TPR", i)]][match(rownames(df), rownames(data))]
    }
  }

  competitors <- setdiff(1:max(brand_lookup$brand_id), brand_num)
  predictors <- c(
    "log_price", "feature", "display", "tpr",
    paste0("log_price", competitors),
    paste0("feature", competitors),
    paste0("display", competitors),
    paste0("tpr", competitors)
  )

  df <- df %>% filter(if_all(all_of(predictors), ~ !is.na(.)))

  model <- lm(
    as.formula(paste("log_sales ~", paste(predictors, collapse = " + "))),
    data = df
  )

  step_model <- tryCatch({
    step(model, direction = "backward", trace = 0)
  }, error = function(e) {
    message("Stepwise failed for Brand ", brand_num, ": ", e$message)
    model
  })

  tidy(step_model) %>%
    mutate(
      Brand = paste0("Brand ", brand_num),
      r_squared = summary(step_model)$r.squared
    )
}

all_brand_results <- purrr::map_dfr(1:max(brand_lookup$brand_id), sales_model)

head(all_brand_results)
```

# Plotting Competitive Map

## Setting up table with results for competitive map plot

```{r}
summary_table_sales <- all_brand_results %>%
  mutate(term_type = case_when(
    term == "log_price" ~ "own_price",
    grepl("^log_price\\d", term) ~ "cross_price",
    grepl("^feature\\d", term) | grepl("^display\\d", term) | grepl("^tpr\\d", term) ~ "cross_promo",
    TRUE ~ "other"
  )) %>%
  group_by(Brand) %>%
  summarize(
    OwnPriceElasticity = if (any(term_type == "own_price")) estimate[term_type == "own_price"][1] else 0,
    MaxCrossPriceEffect = if (any(term_type == "cross_price")) max(estimate[term_type == "cross_price"], na.rm = TRUE) else 0,
    MaxpromoEffect = if (any(term_type == "cross_promo")) max(abs(estimate[term_type == "cross_promo"]), na.rm = TRUE) else 0,
    R2 = max(r_squared, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(brand_lookup %>% mutate(Brand = paste0("Brand ", brand_id)), by = "Brand")

print(summary_table_sales)
```

## Competitive Map Plot

```{r, fig.height=6, fig.width=8}
ggplot(summary_table_sales, aes(x = OwnPriceElasticity, y = MaxCrossPriceEffect, label = DESCRIPTION)) +
  geom_point(aes(size = R2, color = MaxpromoEffect), alpha = 0.8) +
  ggrepel::geom_text_repel(max.overlaps = 100, size = 3.5, box.padding = 0.5) +
  scale_color_gradient2(low = "blue", mid = "gray", high = "red", midpoint = 0, name = "Promotion Sensitivity") +
  scale_size_continuous(name = "Model R²") +
  scale_x_continuous(
    limits = c(-15, 15),
    breaks = seq(-15, 15, by = 3)
  ) +
  labs(
    title = "Mouthwash Category: Competitive Map",
    x = "Own Price Elasticity",
    y = "Cross-Price Elasticity"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

# Additional Tables for Insights

## Creating elasticity matrixes for each type of predictor

```{r}
extract_effect_matrix <- function(all_results, brand_lookup, prefix) {
  brand_names <- brand_lookup$DESCRIPTION
  n_brands <- length(brand_names)
  mat <- matrix(NA, nrow = n_brands, ncol = n_brands)
  rownames(mat) <- brand_names  
  colnames(mat) <- brand_names  

  for (j in 1:n_brands) {
    brand_results <- all_results %>% filter(Brand == paste0("Brand ", j)) 
    own_val <- brand_results %>% filter(term == prefix) %>% pull(estimate)
    mat[j, j] <- ifelse(length(own_val) == 1, own_val, 0)

    for (i in setdiff(1:n_brands, j)) {
      term_i <- paste0(prefix, i)
      cross_val <- brand_results %>% filter(term == term_i) %>% pull(estimate)
      mat[i, j] <- ifelse(length(cross_val) == 1, cross_val, 0)
    }
  }

  round(as.data.frame(mat), 3)
}

price_matrix   <- extract_effect_matrix(all_brand_results, brand_lookup, "log_price")
feature_matrix <- extract_effect_matrix(all_brand_results, brand_lookup, "feature")
display_matrix <- extract_effect_matrix(all_brand_results, brand_lookup, "display")
tpr_matrix     <- extract_effect_matrix(all_brand_results, brand_lookup, "tpr")

# write.csv(price_matrix, "price_matrix.csv")
# write.csv(feature_matrix, "feature_matrix.csv")
# write.csv(display_matrix, "display_matrix.csv")
# write.csv(tpr_matrix, "tpr_matrix.csv")
```

## Creating a long dataframe of each combination to evaluate all variables at once

```{r}
matrix_to_long <- function(effect_df, effect_name) {
  effect_df %>%
    rownames_to_column(var = "Responding_Brand") %>%
    pivot_longer(-Responding_Brand, names_to = "Actioning_Brand", values_to = effect_name) %>%
    select(Actioning_Brand, Responding_Brand, all_of(effect_name))
}

price_long   <- matrix_to_long(price_matrix, "Price_Effect")
feature_long <- matrix_to_long(feature_matrix, "Feature_Effect")
display_long <- matrix_to_long(display_matrix, "Display_Effect")
tpr_long     <- matrix_to_long(tpr_matrix, "TPR_Effect")

brand_names <- brand_lookup$DESCRIPTION
full_combinations <- expand.grid(
  Actioning_Brand = brand_names,
  Responding_Brand = brand_names,
  stringsAsFactors = FALSE
)

combined_effects_df <- full_combinations %>%
  left_join(price_long,   by = c("Actioning_Brand", "Responding_Brand")) %>%
  left_join(feature_long, by = c("Actioning_Brand", "Responding_Brand")) %>%
  left_join(display_long, by = c("Actioning_Brand", "Responding_Brand")) %>%
  left_join(tpr_long,     by = c("Actioning_Brand", "Responding_Brand")) %>%
  mutate(across(ends_with("_Effect"), ~ replace_na(., 0)))

# write.csv(combined_effects_df, "combined_effects_df.csv")
```